### 2012-10-27 thesis/tex/4_visualformalismsforsafetyconstraints.tex
correction, v.1.

\chapter{Visual formalisms}
\label{chap:visualformalisms}

In this chapter we introduce two different approaches of visual formalisms and later state the final desing of the visual language.
One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points. A first approach, so called template constraints and its pros and cons are described in section~\ref{sec:templateconstraints}. Subsequently an improved design is presented, which tends to a more abstract direction.

\section{Template constraints}
\label{sec:templateconstraints}

%It is not a trivial issue to find a
What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Leaded by this question, we developed a visual language based on templates. These templates make different propositions about states and their relations. The user can choose from several graphical constraint types and customize them by parameterizing state variables. So far there are two different kinds of template constraints:

@@

	The idea behind this constraint is, that propositions about temporal dependencies can be made. Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled. Example of a constraint: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z). Going to bed without brushing teeth after meal should never happen.
	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.

	This constraint type enforces at least one particular event to happen after a spezific precondition.
	Example of a constraint: Every time a person enters a supermarket (x) he will eventually decide to buy 
	%a product or not to buy it.
	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can't avoid both possibilities.
	Nitpicking readers might note the person could also steel something without buying it, and thus not buy anything but leave the shop with full hands. Of course this is true; exactly in this case our constraint would signal an undesirable "program" behaviour with its invalidity.
\end{enumerate}

Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like. Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot. Furthermore the state labels should sometimes be replaced by the screen displays of the regarding states. All parameters of the template constraints can be customized by just clicking on one robot suface. In the subsequently appearing popup window (figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.7]{templateconstraint_dropdown}
  \caption{Dropdown menu for spezifying template parameter.}
  \label{fig:templateconstraint_dropdown}
\end{figure}

This visual language was presented to potential users for evaluation and feedbak. Though most of them named it a good concept, soon it turned out that there are some problems and disagreements. The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
A try to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
But unfortunately even the textual eplanations don't seem to be a satisfying solution. For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.

However, there is a second lack with the template constraints.
Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can't be described. Not even the check of a state being visited never or at least once is possible. This lack of expressiveness appears to be too big and for this reason the approach of template constraints got dismissed.

@@

\section{Operator constraints}
\label{sec:operatorconstraints}

With our second approach we try to gain more expressiveness and uniqueness for the visual language. For this we decided to decrease the abstraction level and design a visual formalism for linear temporal logic (LTL, see~section~\ref{sec:modelcheckingandltl}). Since the visual constraints can be directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness. If it is considered as easy to use, two birds get killed with one stone: Non-experts can use it as well as experts.

@@

%The presented visual language is designed to express constraints for state machines.
To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below. In addition the visual language should be capable of expressing constraints about future steps, both (h) any future state and (g) the next state, that (f) events should always happen, and that (i) a property must be true until some future event. Like in Del Bimbo et al.'s formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy. However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
Each block represents either an operator or a proposition and has its specified semantics. Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.

The visual language has support for eight operator types and one proposition as shown in figure~\ref{fig:operators}.

\begin{description}
	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided. Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state. Other types such as numeric or string equations may be added but are less important for our healthcare subject.
	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
	%\Box \Diamond \medcircle
	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true. In addition eventually there has to be a future state with $\psi$ being true.
\end{description}

>>>>

\chapter{Visual formalisms for safety constraints}
\label{chap:visualformalismsforsafetyconstraints}

This chapter introduces two different approaches of visual formalisms and later states the final desing of the visual language.
One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points. A first approach, so called template constraints, and its pros and cons are described in section~\ref{sec:templateconstraintformalism}. Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.

\section{Template constraint formalism}
\label{sec:templateconstraintformalism}

%It is not a trivial issue to find a
What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Led by this question, a visual language based on templates has been developed. These templates make different propositions about states and their relations. The user can choose from several graphical constraint types and customize them by parameterizing state variables. So far there are two different kinds of template constraints:

@@

	The idea behind this constraint is that propositions about temporal dependencies can be made. Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled. Example of a constraint: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z). Going to bed without brushing teeth after meal should never happen.
	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.

	This constraint type enforces at least one particular event to happen after a specific precondition.
	An example of such a constraint would be: Every time a person enters a supermarket (x) he will eventually decide to buy 
	%a product or not to buy it.
	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can't avoid both possibilities.
	%Nitpicking readers might note 
	Of course the person could also steal something without buying it, and thus not buy anything but leave the shop with full hands. Of course this is true; In fact, exactly in this case the constraint would signal an undesirable ``program'' behaviour with its invalidity.
\end{enumerate}

Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like. Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot. Furthermore the state labels should sometimes be replaced by the screen displays of the regarding states. All parameters of the template constraints can be directly customized by the robot surface symbols. In the subsequently appearing popup window (see figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.7]{templateconstraint_dropdown}
  \caption{Dropdown menu for specifying template parameter.}
  \label{fig:templateconstraint_dropdown}
\end{figure}

This visual language was presented to potential users for evaluation and feedback. Though most of them named it a good concept, soon it turned out that there are some problems and disagreements. The presented constraints were interpreted differently, the semantic seemed to be unclear and ambiguous.
An attempt to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
But unfortunately even the textual explanations do not seem to be a satisfying solution. For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.

However, there is a second problem with the template constraints.
Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can't be described. Not even the check of a state being visited never or at least once is possible. This lack of expressiveness appears to be too big and for this reason the approach of template constraints got dismissed.

@@

weitere constraint templates...

\section{Operator constraint formalism}
\label{sec:operatorconstraintformalism}

With our second approach we try to gain more expressiveness and uniqueness for the visual language. For this we decided to decrease the abstraction level and design a visual formalism for linear temporal logic (LTL, see~section~\ref{sec:modelcheckingandltl}). Since the visual constraints can be directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness. If it is considered as easy to use,
%two birds get killed with one stone: N
non-experts can use it as well as experts.

@@

%The presented visual language is designed to express constraints for state machines.
To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below.

\begin{description}
	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided. Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state. Other types such as numeric or string equations may be added but are less important for our healthcare subject.

In addition the visual language should be capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i).

	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
	%\Box \Diamond \medcircle
	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true. In addition eventually there has to be a future state with $\psi$ being true.
\end{description}

Like the formula representation introduced by Del Bimbo et al.~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy. However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
Each block represents either an operator or a proposition and has its specified semantics. Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.

The visual language has support for eight operator types and one proposition. A suggestion for their graphical representation is shown in figure~\ref{fig:operators}.

### 2012-10-27 thesis/tex/5_automatedgenerationofsafetyconstraints.tex
correction, v.1.

\chapter{Automated constraint generation}
\label{chap:automatedconstraintgeneration}

As already mentioned before, it might be difficult for developers to find all reasonable safety constraints by hand which are needed for full safety covery. Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
To help prevent this to happen, we introduce the concept of automated constraint generation for state machines. Once a program reaches a complete condition, constraint suggestions which are valid on the current program can be computed fully automatically. On the one hand they may help the developer identifying reasonable constraints he didn't think about as well as contradictions between program and specification, eg. if a computed constraint doesn't make sense at all. On the other hand all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs, automated generation and suggestion of constraints might be helpful. But it's helpful only if the suggested constraints are relevant and reasonable.

@@

Whether suggested constraints are relevant and reasonable
%This however
depends on the domain and use case. It has to be mentioned that our approach of finding an algorithm for constraint finding presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
Regarding the important constraints postulated in chapter~\ref{chap:goals}, we found out that all all the relevant states fall into one of the following categories:

@@

In order to compute the required constraints automatically, and possibly additional ones, we deal with subgraphs defined as follows:
%Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

@@

@@

A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine.

@@

If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):

@@

These five constraints apply to each subgraph matching the conditions mentioned above. Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state. Formula~\ref{eq:second} ensures that there will be always at least one $\gamma$-state be visited after each visit of an $\alpha$-state. The quite similar but stricter formula~\ref{eq:third} demands additionally a visit of at least one $\gamma$-state before the $\beta$-state can be reached. Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states. The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

All formulas can be translated to the visual formalism presented in section~\ref{sec:operatorconstraints} and shown to the user. The appropriate LTL formulas for the constraints postulated in chapter~\ref{chap:goals} relate to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:

\begin{equation}
  \models \Box \Diamond \textnormal{'Polling'}
\end{equation}

\begin{equation}
	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
\end{equation}
%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.

>>>>

\chapter{Automated generation of safety constraints}
\label{chap:automatedgenerationofsafetyconstraints}

Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications. In order to discover whether a program meets its safety requirements, its behaviour is checked by the constraints. This happens mostly as a last check before software gets released. 
But regarding updating and maintenace work, often further changes are made to the software even after release.
%But even after possible changes to the software in the context of updating and maintenance work,
However, safety constraints can be applied again and assure the adherence of all properties. Safety constraints thus are also suitable for sustaining consistency during the different phases of software life. In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that. The more constraints are defined the the better can be the reliability.
But as already mentioned before, it might be difficult for developers to find all reasonable %safety
constraints by hand. % which are needed for full safety covery.
In this case it can happen that a prior correct software version becomes faulty over time because of minor adjustments once in a while which are not covered by constraints.
%To help prevent this to happen

As a solution to this problem, the concept of automated constraint generation for state machines is proposed. Once the initial implementation of a program is finished, its behaviour can be analysed. Sgnificant parts of it can be determined automatically and put into constraint representation which are valid on the current program of course.
%constraint suggestions which are valid on the current program can be computed fully automatically. 
On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg. if a computed constraint does not make sense at all.
On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs and to simplify maintainability, automated generation and suggestion of constraints might be helpful. But it's helpful only if the suggested constraints are relevant and reasonable.

@@

Whether suggested constraints are relevant and reasonable
%This however
depends on the domain and use case. It has to be mentioned that the approach of finding an algorithm for constraint generation presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
Regarding the important safety constraints postulated in chapter~\ref{chap:goals}, all the relevant states seem to fall into one of the following categories:

@@

%All associated states which match these properties can be combined to a 
In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined. Such a group of states is named subgrap from now on. It is defined as follows:
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

@@

A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible subgraphs of this kind can be detected within a state machine. An algorithm how subgraphs can be computed programmatically is discussed later in section~\ref{sec:prototype:automatedconstraintgeneration}.

@@

@@

For the healthcare applications, all important states which concern safety constraints belong to subgraphs. Thus, several safety properties can be formulated based on the structure of subgraphs.
For each subgraph the following valid LTL formulas can be derived if it contains neither inner (infinite) loops nor final states (states having no outgoing transition). In the following $\alpha$ is proposition for ``$\alpha$ is current state'' and similarly $\beta$ and $\gamma$:

@@

These five constraints apply to each subgraph matching the conditions mentioned above. Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state. Formula~\ref{eq:second} ensures that there will always at least one $\gamma$-state be eventually visited after each visit of an $\alpha$-state. The quite similar but stricter formula~\ref{eq:third} additionally demands a visit of at least one $\gamma$-state before the $\beta$-state can be reached. Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states. The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

All formulas can be translated to the visual formalism presented in section~\ref{sec:operatorconstraintformalism} and shown to the user. The appropriate LTL formulas for the constraints postulated in chapter~\ref{chap:goals} relate to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:

\begin{equation} \label{eq:sixth}
  \models \Box \Diamond \textnormal{'Polling'}
\end{equation}

\begin{equation} \label{eq:seventh}
	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
\end{equation}
%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.

Formula~\ref{eq:sixth} ensures that state ``Polling'' will always eventually be visited again, and formula~\ref{eq:seventh} specifies that either state ``Well done!'' or state ``Staff notified'' will be eventuelly visited after each visit of ``Not taken yet''.
